---
title: "`purrr` Tutorial"
author: "Dave Bosworth"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
library(tidyverse)
library(here)
```

# Simple Examples of `map()`

Create a data frame to apply functions to:

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

glimpse(df)

df %>% map(5)
```

Use `map()` to calculate summary statistics on each column in `df`

```{r}
map(df, mean)

# With pipe
df %>% map(mean)

# Other summary statistics
df %>% map(median)
df %>% map(sd)

# map() returns a list
df %>% map(mean) %>% class()
```

What if I want `map()` to return a numeric vector instead?

```{r}
df %>% map_dbl(mean)
df %>% map_dbl(mean) %>% class()
```

# Intermediate Example

```{r}
mtcars %>% 
  split(.$cyl) %>% 
  map(~ lm(mpg ~ wt, data = .)) %>% 
  map(summary) %>% 
  map_dbl("r.squared")
```

# More than one input

Can use `map2()` if you have more than one input to the function

```{r}
means <- 1:4
sds <- 1:4

map2(means, sds, rnorm, n = 5)
map2(means, sds, ~ rnorm(mean = .x, sd = .y, n = 5))
```

# Practical Examples

Using `map()` to import multiple .csv files at once

```{r}
fp_chla <- dir(here("data"), full.names = TRUE)

ls_chla <- map(fp_chla, read_csv, col_types = cols(.default = "c"))

walk(ls_chla, glimpse)

df_chla <- list_rbind(ls_chla)

df_chla_c <- df_chla %>% 
  mutate(
    time = parse_date_time(time, orders = c("Ymd T", "mdY R"), tz = "Etc/GMT+8"),
    value = as.numeric(value)
  ) %>% 
  drop_na(value) %>% 
  filter(qaqc_flag_id != "X") %>% 
  select(
    Station = station,
    DateTime = time,
    Chla = value
  )
```

Using a nested dataframe and `map()` to create plots for each Station

```{r}
plot_chla <- function(df, station) {
  df %>% ggplot(aes(x = DateTime, y = Chla)) + 
    geom_point(color = "darkcyan", alpha = 0.4) +
    geom_smooth(color = "black", se = FALSE) +
    theme_bw() +
    ggtitle(station)
}

ndf_chla <- df_chla_c %>% nest(data = -Station)

ndf_chla_plt <- ndf_chla %>% mutate(plt = map2(data, Station, plot_chla))

ndf_chla_plt$plt[[1]]
```

Export plots as .jpg files

```{r}
walk2(
  ndf_chla_plt$Station,
  ndf_chla_plt$plt,
  ~ ggsave(
    filename = paste0(.x, ".jpg"),
    plot = .y,
    path = here("plots")
  )
)
```

